---
title: "Expanding linear regression models"
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(broom)
library(dplyr)
library(tidyr)
library(faraway)
library(patchwork)
library(readxl)
```

### Goals
So far we have looked at the simplest case of a regression model, in which we have one continuous X variable and one continuous Y variable.  Now we want to build on that base to include

  - categorical predictor variables
  - models with more variables

## Categorical predictors
If we have a predictor variable that is categorical rather than continuous we can still use a linear model to describe the relationship between x and y.
$$y= \beta_{0} + \beta_{1}*x$$
Consider a model where x is exposure to treatment and is coded 0= no treatment 1= treatment and y equals systolic blood pressure.

$$blood pressure= \beta_{0} + \beta_{1}*treatment$$

then the average expected value for blood pressure among those who had treatment would be $\beta_{0} + \beta_{1}$  and the expected value for blood pressure among those who did not have treatment would be $\beta_{0}$ (if treatment=0  the $\beta_{1}$ would be multiplied by 0)

## Categorical predictors
```{r}
set.seed(241)
# here I am generating a set of x that has 200 observations and has a probability of 50%
X <- (rbinom(200, size=1, prob=.5))
# generating a normally distributed error
e<-rnorm(200, mean=0, sd=.8)
## and creating Y 
Y<-140 - X*15+e

simbp <-data.frame(cbind(X,Y))
ex1<-ggplot(simbp, aes(x=X, y=Y))+
  geom_point()+
  theme_minimal(base_size=15)
```

### graphically
```{r, echo=FALSE , fig.dim(4,2.5), out.width="60%" }
ex1
```

### graphically
```{r, echo=FALSE, out.width="60%"}
 ggplot(aes(x=Y), data=simbp) + 
  geom_histogram(aes(fill=X),bins=15) + 
  theme_minimal() +
  facet_wrap(~X, nrow=2) +
  labs(title = "Systolic bp by treatment", x = "treatment")
```

### linear model
```{r}
lm_bp<-lm(Y ~ X, data=simbp)
tidy(lm_bp)
glance(lm_bp)
```


### linear model vs t test
```{r}
t.test(Y ~ X, data=simbp)
```

### binary 0/1 vs 1/2 coding
```{r}
simbp2<-simbp%>%mutate(X2=X+1)
lm_bp2<-lm(Y ~ X2, data=simbp2)
coefficients(lm_bp)
coefficients(lm_bp2)
```

How does this change the equation?  our interpretation?  Why do we generally use 0/1?

## Creating a binary predictor from a continuous predictor
Let's imagine that we are studying birthweights by gestational age.  Both are collected continuously, so we might graph these data as a scatterplot.  Here I will simulate a dataset based on the distribution of Canadian Female Singleton births (reference Michael S. Kramer et al. A New and Improved Population-based Canadian Reference for Birth Weight for Gestational Age.)[http://www.pediatrics.org/cgi/content/full/108/2/e35] and graph the results:

```{r, echo=FALSE}
refs<-read_excel("bwt_refs.xlsx")
GA<-rep(22:43, 100)
gas<-data.frame(GA)
brths<-left_join(gas, refs, by="GA")%>%mutate(bwt=rnorm(2200,mean=Mean, sd=SD), premature=ifelse(GA<37, 1,0))
```

### Graph
```{r, echo=FALSE, out.width="80%"}
ggplot(data=brths, aes(x=GA, y=bwt))+
  geom_point()+
  theme_minimal(base_size=15)+
  labs(title="Birthweight by Gestational age", x="Gestational age in weeks", y="Birthweight in grams", caption="Canadian Female Singleton births")
```

### Linear Model for birthweight by GA
```{r}
bwt_ga_cont<-lm(bwt~ GA, data=brths)
tidy(bwt_ga_cont)
glance(bwt_ga_cont)
```

### Model fit
```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="80%"}

augmented_dat1 <- augment(bwt_ga_cont)

## Fitted model
plot1 <- ggplot(augmented_dat1, aes(GA, bwt)) +
  geom_smooth(method = "lm", se = F) +
  geom_point() +
  geom_segment(aes(xend = GA, yend = .fitted), lty = 2) +
  theme_minimal(base_size = 15) +
  labs(title = "(a) Scatter plot") 

# QQ plot
plot2 <- ggplot(augmented_dat1, aes(sample = .resid)) + 
  geom_qq() + 
  geom_qq_line() +
  theme_minimal(base_size = 15) +
  labs(y = "Residuals", x = "Theoretical quantiles", title = "(b) QQplot") 

## Fitted vs. residuals
plot3 <-ggplot(augmented_dat1, aes(y = .resid, x = .fitted)) + 
  geom_point() + 
  theme_minimal(base_size = 15) +
  geom_hline(aes(yintercept = 0)) +
  labs(y = "Residuals", x = "Fitted values", title = "(c) Fitted vs. residuals") 

## Amount explained
reshape <- augmented_dat1 %>% select(.resid, bwt) %>%
  gather(key = "type", value = "value", bwt, .resid)

plot4 <- ggplot(reshape, aes(y = value)) + 
  geom_boxplot(aes(fill = type)) +
  labs(title = "(d) Amount explained") +
  theme_minimal(base_size = 15) 

library(patchwork)
plot1 + plot2 + plot3 + plot4 + plot_layout(nrow = 2) + 
  plot_annotation(caption = "Continuous vs Continuous")
```

### What happens if we categorize? 
We often look at birth outcomes by premature status.  If I create a 0/1 indicator variable for gestational age less than 37 weeks  the distribution looks like this:
```{r, echo=FALSE, out.width="60%"}
ggplot(aes(x=bwt), data=brths) + 
  geom_histogram(aes(fill=as.factor(premature)),bins=15) + 
  theme_minimal() +
  facet_wrap(~premature, nrow=2) +
  labs(title = "Birthweight by prematurity status", x = "Birthweight in grams", fill="Premature yes/no")
```

### linear model 
```{r, echo=FALSE}
bwt_ga_bin<-lm(bwt~ premature, data=brths)
tidy(bwt_ga_bin)
glance(bwt_ga_bin)
```

### t-test
```{r, echo=FALSE}
t.test(bwt ~ premature, data=brths)
```


### Model fit for binary predictor
```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="80%"}

augmented_dat1 <- augment(bwt_ga_bin)

## Fitted model
plot1 <- ggplot(augmented_dat1, aes(premature, bwt)) +
  geom_smooth(method = "lm", se = F) +
  geom_point() +
  geom_segment(aes(xend = premature, yend = .fitted), lty = 2) +
  theme_minimal(base_size = 15) +
  labs(title = "(a) Scatter plot") 

# QQ plot
plot2 <- ggplot(augmented_dat1, aes(sample = .resid)) + 
  geom_qq() + 
  geom_qq_line() +
  theme_minimal(base_size = 15) +
  labs(y = "Residuals", x = "Theoretical quantiles", title = "(b) QQplot") 

## Fitted vs. residuals
plot3 <-ggplot(augmented_dat1, aes(y = .resid, x = .fitted)) + 
  geom_point() + 
  theme_minimal(base_size = 15) +
  geom_hline(aes(yintercept = 0)) +
  labs(y = "Residuals", x = "Fitted values", title = "(c) Fitted vs. residuals") 

## Amount explained
reshape <- augmented_dat1 %>% select(.resid, bwt) %>%
  gather(key = "type", value = "value", bwt, .resid)

plot4 <- ggplot(reshape, aes(y = value)) + 
  geom_boxplot(aes(fill = type)) +
  labs(title = "(d) Amount explained") +
  theme_minimal(base_size = 15) 

library(patchwork)
plot1 + plot2 + plot3 + plot4 + plot_layout(nrow = 2) + 
  plot_annotation(caption = "binary predictor")
```


## Categorical predictors with more than 2 levels

### Adding categories
Remember that a coefficient in a linear model represents the change in Y related to a one unit change in X, so how do we handle this for categories?  Even in ordinal variables - those that have an inherent order - we generally do not expect the difference between each category to be the same.  For nominal variable we would be even less comfortable with that structure....so


### Adding categories
As long as the categories are mutually exclusive and exhaustive we can create a series of "dummy" variables coded 0/1 that represent the categories.  

$$y= \beta_{0} + \beta_{1}*category_1 + \beta_{2}*category_2 \cdot \cdot\cdot + \beta_{j}*category_{k-1}$$
where k is the number of categories. Why do we have 1 fewer terms than categories?
Let's look at an example...

### Example:  Butterfat
This dataset contains data on the average butterfat content of milk from random samples of diary cattle.  Data came from Canadian records of pure-bred dairy cattle 

```{r}
data(butterfat, package="faraway")
str(butterfat)
```

### Example:  Butterfat
We can look at the means by group here
```{r}
butterfat%>% 
  group_by(Breed) %>% 
  summarize(mean_fat = mean(Butterfat, na.rm=TRUE))
```

### Example:  Butterfat
and graph them here:
```{r, echo=FALSE, out.width="70%"}
ex2<-ggplot(butterfat, aes(x=Breed, y=Butterfat))+
  geom_point()+
  theme_minimal(base_size=15)
ex2
```

### Example:  Butterfat
What does the model output look like?
```{r}
lm_fat<-lm(Butterfat ~ Breed, data=butterfat)
tidy(lm_fat)
```

**where is the expected value for Ayrshire?**

### choosing a different reference category
```{r}
butterfat_relevel <- within(butterfat, Breed <- relevel(Breed, ref = "Holstein-Fresian"))

lm_fat2<-lm(Butterfat ~ Breed, data=butterfat_relevel)

coefficients(lm_fat)
coefficients(lm_fat2)

```

## models with more than one predictor variable

### Regression line with multiple x variables
In the simple case our regression line was:
$$y= \beta_{0} + \beta_{1}*x$$
in the case of multiple variables we can write this more generally as:

$$y= \beta_{0} + \beta_{1}*x_{1} +\beta_{2}*x_{2} + \cdot \cdot \cdot +\beta_{p}*x_{p}$$

where $p$ is the number of x variables included in the model.

### interpreting the coefficients 
In a simple model, we interpreted the $\beta$ or slope coefficient as the change in the predicted or "expected" value of y associated with a one unit change in x.

When there are multiple x variables, the interpretation is that the coefficient represents the change in the expected value of y associated with a one unit change in x, **holding all other factors in the model constant** or "controlling for" the other factors in the model.


### Example: Butterfat 
Let's say we also suspect that there is a difference in butterfat content as cows reach maturity
```{r}
butterfat%>% 
  group_by(Age) %>% 
  summarize(mean_fat = mean(Butterfat, na.rm=TRUE))
```

### Example:  Butterfat
Including age in the visualization by breed
```{r, echo=FALSE, out.width="70%"}
ex2<-ggplot(butterfat, aes(x=Breed, y=Butterfat, color=Age))+
  geom_point()+
  theme_minimal(base_size=15)
ex2
```

### Butterfat model with both factors
```{r}
lm_fat3<-lm(Butterfat ~ Breed + Age, data=butterfat_relevel)
tidy(lm_fat3)
```

### Butterfat model with both factors
$$Butterfat= \beta_{0} + \beta_{1}*Breed_{1} +\beta_{2}*Breed_{2} + \beta_{3}*Breed_{3}+ \beta_{4}*Breed_{4}+ \beta_{5}*Matureage$$

$$= 3.62 + 0.39*Ayrshire +0.7690*Canadian + 1.28*Guernsey+ 1.62*Jersey+ 0.10*Matureage$$
What is $\beta_{0}$ in this model?


### comparing any 2 groups
What if we are interested in the difference in butterfat between milk from a young Ayrshire cow and a mature Guernsey?

Young Ayrshire:
$$E[Butterfat |Breed= Ayrshire, Age=young]= 3.62 + 0.39*Ayrshire = 4.01$$

Mature Guernsey:
$$E[Butterfat | Breed=Guernsey, Age=Mature]= 3.62 +1.28*Guernsey+ 0.10*Mature = 5$$

$$E[Butterfat |Breed= Ayrshire, Age=young] - E[Butterfat | Breed=Guernsey, Age=Mature] $$
$$= 4.01 - 5 = -0.99$$
So the difference is 0.99 less butterfat among young Ayrshire cows compared to mature Guernseys

### testing for variables with multiple categories
When we are looking at variables that are continuous or binary [0,1] the t-test values for each $\beta$ give us a sense of the importance of each variable.  

Why is this a problem in a variable with multiple categories?

```{r, echo=FALSE}
tidy(lm_fat3)
```

### comparing models f-test
We will use the `anova()` function in r to compare a model that includes Breed to one that does not
```{r}
lm_fat4<-lm(Butterfat ~  Age, data=butterfat_relevel)

anova(lm_fat4, lm_fat3)
```


### comparing models R-squared
We might also be interested in comparing the amount of variability explained by our two models:

```{r}
glance(lm_fat4)
glance(lm_fat3)
```
